Appendix 4: Dynamic species distribution model comparison between INLA, inlabru, and sdmTMB

### 2022-01-08 doc/appendix-binomial-inla.Rmd
refine order and minor edits to inla appendix

@@

m_inla <- inla(formula,
  data = inla.stack.data(sdat),
  control.predictor = list(compute = TRUE, A = inla.stack.A(sdat)),
  family = "binomial",
  control.family = list(link = "logit"),
  control.fixed = list(expand.factor.strategy = "inla", mean = 0, prec = 1 / (100 * 100)),
  control.inla = list(int.strategy = "eb", strategy = "gaussian")
)
tictoc::toc() # stop timing

summary(m_inla)
inla_coefs <- m_inla$summary.fixed[, c("mean", "sd")]
```

@@

```{r}
summary(m_bru)

m_inla$summary.hyperpar[, c("mean", "sd")]
m_bru$summary.hyperpar[, c("mean", "sd")]

bru_coefs <- m_bru$summary.random$fac[, c("ID", "mean", "sd")]
m_inla$summary.fixed[,c("mean", "sd")]
bru_coefs
```

@@

```{r sdmtmb-coefs}
print(m)
tidy(m, "ran_pars", conf.int = TRUE)
sdmTMB_coefs <- tidy(m, effects = "fixed", conf.int = TRUE)
```

@@

Comparing the fixed effects between `sdmTMB()` and `inla()` shows the estimates are very similar:

```{r coef-plot, out.width="5.5in"}
inla_coefs <- m_inla$summary.fixed[, c("0.025quant", "mean", "0.975quant")] %>% 
  rename(estimate = mean, conf.low = `0.025quant`, conf.high = `0.975quant`) %>% 
  mutate(term = unique(sdmTMB_coefs$term), model = "INLA")

@@

We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the sdmTMB package to map occurrence probabilities in each year in space.

@@

```{r maps, cache=FALSE, out.width="6in"}
ggplot(pred) +
  geom_raster(aes(X, Y, fill = plogis(est))) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

@@

```{r inla-pred, out.width="6in"}
idx <- inla.stack.index(sdat, "stdata")$data
d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
ggplot(d) +
  geom_point(aes(X, Y, colour = inla_pred), size = 2) +
  facet_wrap(~year) +
  scale_colour_viridis_c() +
  coord_fixed()
```

Or, we can make the prediction with inlabru. To do that, we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and use the inverse-link transformation `plogis()`:

```{r inlabru-pred, out.width="6in"}
nd_bru <- nd
nd_bru <- dplyr::left_join(nd_bru, yr_lu, by = "year")
sp::coordinates(nd_bru) <- c("X", "Y")
nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))
ggplot(as.data.frame(pred_bru)) +
  geom_raster(aes(X, Y, fill = mean)) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

>>>>

# Background

@@

m_inla <- inla(formula,
  data = inla.stack.data(sdat),
  control.predictor = list(compute = TRUE, A = inla.stack.A(sdat)),
  family = "binomial",
  control.family = list(link = "logit"),
  control.fixed = list(expand.factor.strategy = "inla", mean = 0, prec = 1 / (100 * 100)),
  control.inla = list(int.strategy = "eb", strategy = "gaussian")
)
tictoc::toc() # stop timing
```

```{r}
summary(m_inla)
```

@@

```{r}
summary(m_bru)
```

```{r}
m_bru$summary.random$fac[, c("ID", "mean", "sd", "0.025quant", "0.975quant")]
```

@@

```{r sdmtmb-m}
print(m)
```

We can most easily extract the fixed-effect estimates and confidence intervals using smdTMB's `tidy()` function.

```{r sdmtmb-coefs}
sdmTMB_coefs <- tidy(m, effects = "fixed", conf.int = TRUE)
sdmTMB_coefs
```

\clearpage

@@

The estimates for the spatial parameters are very similar between the INLA-based models and `sdmTMB()`:

```{r}
# INLA
m_inla$summary.hyperpar[, c("mean", "sd", "0.025quant", "0.975quant")]

# inlabru
m_bru$summary.hyperpar[, c("mean", "sd", "0.025quant", "0.975quant")]

# sdmTMB
tidy(m, "ran_pars", conf.int = TRUE)
```

Note that the standard errors for variance terms are not in natural and not log space and so are a very rough approximation and therefore not reported by the `tidy()` function.

Comparing the fixed effects shows the estimates from all three models are also very similar:

```{r coef-plot, fig.asp=0.4, out.width="85%"}
inla_coefs <- m_inla$summary.fixed[, c("0.025quant", "mean", "0.975quant")] %>% 
  rename(estimate = mean, conf.low = `0.025quant`, conf.high = `0.975quant`) %>% 
  mutate(term = unique(sdmTMB_coefs$term), model = "INLA")

@@

We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the sdmTMB package to map occurrence probabilities in each year in space. 
The resulting values are in link space and require the appropriate inverse-link transformation, in this case `plogis()`, to produce occurrence probabilities.

@@

```{r maps, cache=FALSE, out.width="5.5in"}
ggplot(pred) +
  geom_raster(aes(X, Y, fill = plogis(est))) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

We can make similar predictions with inlabru using only a few additional steps: we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and the appropriate inverse-link transformation:

```{r inlabru-pred, out.width="5.5in"}
nd_bru <- nd
nd_bru <- dplyr::left_join(nd_bru, yr_lu, by = "year")
sp::coordinates(nd_bru) <- c("X", "Y")
nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))

ggplot(as.data.frame(pred_bru)) +
  geom_raster(aes(X, Y, fill = mean)) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

@@

```{r inla-pred, out.width="5.5in"}
idx <- inla.stack.index(sdat, "stdata")$data
d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]

ggplot(d) +
  geom_point(aes(X, Y, colour = inla_pred), size = 2) +
  facet_wrap(~year) +
  scale_colour_viridis_c() +
  coord_fixed()
```
